---
title: "Report for psykodentopat"
author: "Hilde Kollsete Gjelberg, Ingrid Eline Holm, and Simen Berrefjord"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a R Markdown report for the group exam in RMED901A for the group psykodentopat. The group consists of Hilde Kollsete Gjelberg, Ingrid Eline Holm, and Simen Berrefjord.

The report includes all the steps we have conducted throughout the course related to the exam data (i.e., exploration, tidying, visualization, and statistical analyses).

## Data exploration


## Tidying of data


## Visualization
Our task was to create plots to answer the following questions: 

**1. Are there any correlated measurements?**

```{r reading data, include=FALSE}
library(tidyverse)
library(here)
library(ggplot2)
library(patchwork)
joined_exam_data <- read_delim(here("data","joined_exam_data_2025-09-09.txt"))

```

```{r warning = FALSE}
joined_exam_data %>%
  select(-patient_id) %>% 
  GGally::ggcorr()
```
  
This correlation matrix demonstrates a strong positive correlation between `swallowPain` and `throatPain`.`BMI` and `age` appears to have a weak positive correlation.  
  
  
**2.  Does the age distribution depend on `treat`?**

```{r ggplot patient age treatment, include=FALSE}
ggplot <- ggplot(joined_exam_data, aes(x = age, fill = treat)) +
  geom_density(alpha = 0.5) +
  theme_gray() +
  labs(title = "Age distribution by treatment",
       x = "Patient age",
       fill = "Treatment:") +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "PuOr")
```
 
```{r, echo=TRUE, eval=FALSE}
ggplot(joined_exam_data, aes(x = age, fill = treat)) +
  geom_density(alpha = 0.5) +
  labs(title = "Age distribution by treatment",
       x = "Patient age",
       fill = "Treatment:")
```

```{r, echo=FALSE}
ggplot
```
  
This density-plot demonstrates that the age distributions in the two treatment groups are similar, and age distribution does not seem to depend on treatment. There may be a mild tendency for Licorice-treated patients to be older on average, based on the higher peak in the 65–75 age range.  

**Does the age distribution of the patients depend on their (`gender`)?**
  
  
```{r ggplot age gender, include=FALSE}
density_age_distribution_by_gender <- ggplot(joined_exam_data, 
                                          aes(x = age, fill = gender)) +
  geom_density(alpha = 0.5) +
  labs(title = "Age distribution by gender", 
       x = "Patient age", 
       fill = "Gender") +
  scale_fill_brewer(palette = "Pastel2")
density_age_distribution_by_gender
```
```{r, echo=TRUE, eval=FALSE}
ggplot(joined_exam_data, aes(x = age, fill = gender)) +
  geom_density(alpha = 0.5) +
  labs(title = "Age distribution by gender", 
       x = "Patient age", 
       fill = "Gender")
```
```{r, echo =FALSE}
density_age_distribution_by_gender
```

This density-plot demonstrates that the age distribution does not strongly depend on gender seeing as the density curves overlap substantially.  
    
    
**Does the preoperative pain change with age of the patients?**
```{r}
ggplot(joined_exam_data, aes(x = age, fill = preOp_pain)) +
  geom_density(alpha = 0.4) +
  labs(
    x = "Age",
    y = "Density"                 
  )
```
  
This density-plot can indicate difference in age in those experiencing preoperative pain and not. However, only two patients in the dataset reported pre-operative pain.The density curve for the "yes" group therefore represents very few observations and should be interpreted with caution. This explains why the curve looks irregular compared to the "no" group. We have to limited datamaterial to answer the question.   



## Statistical analyses


## Closing remarks


